Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

### 2023-08-28 manuscript/manuscript.qmd
Revise Hawaii and Manuscript
Revise Hawaii to reduce HeatPlot to show top 10; revise manuscript based on Cory's comments, update supplement.

Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008]. However, the dramatic increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all. This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.

@@

Recordings from the U.S. East Coast (western North Atlantic Ocean) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey by Northeast Fisheries Science Center and Southeast Fisheries Science Center [@united_states_national_marine_fisheries_service_2016_2016]. AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey areas. The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, Appendix A, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (HTI-96-min, High Tech Inc., Long Beach, MS) which recorded at a 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see @deangelis_description_2018 for more information). The SAtlantic area ranged from Delaware south to central Florida (GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (Reson TC4013, Teledyne Marine, Slangerup, Denmark) which recorded at a 500 kHz sample rate (custom 12 channel SailDAQ soundcard) and were decimated to 192 kHz with a 1 kHz high pass filter (see GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016] for more information). The NAtlantic and SAtlantic recordings were treated as two separate datasets due to differences in hydrophone frequency responses and sensitivities.

@@

Echolocation signals that were identified as frequency-modulated pulses from beaked whales were then linked as PAMGuard "events" by experienced acoustic researchers. Events included all the pulses that were close to each other in time, had similar spectral and waveform characteristics, and were received at consistent bearing angles. For NAtlantic and SAtlantic datasets, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into putative individuals by identification of consecutive pulses along the same bearing angle in the bearing time plot (see @deangelis_using_2017).

Experienced acoustic researchers used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale. Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@r_core_team_r_2022].

@@

For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species. Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than expected (Priors in @fig-natlantic a). The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig. 3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model. Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig. 2).

@@

The best model considered the echolocation pulses, IPI, and environmental features (EC_IPI_ENV Model, Detector Model: *sampsize* = 1, *ntree* = 10,000; Event model: *sampsize* = 5, *ntree* = 10,000) and provided an overall correct classification rate of 91.9% for all six species. Classification scores ranged from a low of 91.2% (BW43) to 100% for Cross Seamount beaked whale (@fig-epacific a). All classification results were greater than expected (Prior in @fig-epacific a). Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs. 5).

@@

Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced acousticians are unable to differentiate species (e.g., long-beaked and short-beaked common dolphins, @rankin_acoustic_2017). This suggests that large increases in sample sizes may improve classification results for Cuvier's beaked whales.

@@

These two examples of small sample sizes highlight a conflict in preferred protocol. Ideally, (1) all [species]{.underline} would be included in a classification model, so as to better represent the local species diversity, (2) all [events]{.underline} would ideally be included in the classification model, so as to better represent the variability found in the area, and (3) only confident 'ground truth' classifications would be considered in the training model. Unfortunately, in species where identity must be determined based on call characteristics (rather than visual confirmation of species identity in the field), it can sometimes be difficult to confidently determine species identity when calls are highly variable and do not include at least one call that provides high confidence of species identity. In the case of beaked whales, we recommend that training data include a high level of confidence for inclusion. This is especially critical for species with small sample sizes. An alternative is to require agreement from multiple analysts. This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified. For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysis did not identify these in the EPacific data presented here).

@@

The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies. Thanks to Liam Mueller-Brennan for their help with data analysis. Sea surface temperature data were provided by JPL under support by the NASA MEaSUREs program. Funding for data collection and analysis were provided by the U.S. Navy, the Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration. The AMAPPS study (NAtlantic/SAtantic) was funded in part by the U.S. Department of the Interior, Bureau of Ocean Energy Management, Environmental Studies Program, Washington, DC, through Inter-Agency Agreement Number M10PG00075 with the National Marine Fisheries Service as the Atlantic Marine Assessment Program for Protected Species. The HICEAS 2017 Survey (Hawaii) was funded by Bureau of Ocean Energy Management under (IAA M17PG00024) and NMFS Pacific Islands Fisheries Science Center. The CCES Survey (EPacific) was funded by Bureau of Ocean Energy Management (No. M17PG00025), U.S. Navy Pacific Fleet Environmental Readiness Division (IAA No. N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center. The manuscript was improved thanks to reviews from XX and XX.

>>>>

Passive acoustic monitoring (PAM) has proven to be a valuable tool for studying populations of marine mammals [@parijs_management_2009]; however, the value of these studies depends on our ability to identify the sources of the sounds we are monitoring. Historically, experienced acousticians have manually identified stereotyped sounds that could reliably be attributed to a given species, based on their spectral or temporal characteristics [@baumann-pickering_species-specific_2013]; [@bittle_review_2013]; [@rankin_source_2005]; [@soldevilla_classification_2008]. However, the substantial increase in the quantity of recordings being made make it impossible for experienced acousticians to manually annotate it all. This has led to an increased need for the development of automated classification routines that can provide accurate species determinations from acoustic recordings.

@@

Recordings from the U.S. East Coast (western North Atlantic Ocean) were collected using towed hydrophone arrays during the 2016 AMAPPS II (Atlantic Marine Assessment Program for Protected Species) survey by Northeast Fisheries Science Center and Southeast Fisheries Science Center [@united_states_national_marine_fisheries_service_2016_2016]. AMAPPS II was subdivided into the northern (NAtlantic) and southern (SAtlantic) survey areas. The NAtlantic study area ranged from Massachusetts south to New Jersey (HB1603, Appendix A, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (HTI-96-min, High Tech Inc., Long Beach, MS), which recorded at a 192 kHz sample rate with a 1 kHz high pass filter (National Instruments USB-6356 A/D card, see @deangelis_description_2018 for more information). The SAtlantic area ranged from Delaware south to central Florida (GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016]) and recordings used in this analysis were collected with two hydrophones (Reson TC4013, Teledyne Marine, Slangerup, Denmark), which recorded at a 500 kHz sample rate (custom 12 channel SailDAQ soundcard) and were decimated to 192 kHz with a 1 kHz high pass filter (see GU1605, Appendix B, [@united_states_national_marine_fisheries_service_2016_2016] for more information). The NAtlantic and SAtlantic recordings were treated as two separate datasets due to differences in hydrophone frequency responses and sensitivities.

@@

Echolocation signals that were identified as frequency-modulated pulses from beaked whales were then linked to PAMGuard "events" by experienced acoustic analysts. Events included all the pulses that were close to each other in time, had similar spectral and waveform characteristics, and were received at consistent bearing angles. For NAtlantic and SAtlantic datasets, acoustic detections of beaked whales using towed hydrophone arrays were further subdivided into putative individuals by identification of consecutive pulses along the same bearing angle in the bearing time plot (see @deangelis_using_2017).

Experienced acoustic analysts used multiple lines of evidence (PAMGuard click detector bearing time display, click waveform display, click spectrum, and wigner plots) to assess the species identity of the acoustic event. When the available characteristics were inconclusive, the species was considered an unidentified beaked whale. Data were saved within the PAMGuard database and binaries for downstream processing using the *PAMpal package* [@sakai_pampal_2021] in R v2022.07.02 [@r_core_team_r_2022].

@@

For the SAtlantic dataset, the most accurate model considered the echolocation pulse metrics and the IPI for an event (EC_IPI Model, Detector Model: *sampsize* = 4, *ntree* = 10,000; Event Model: *sampsize* = 1, *ntree* = 100,000) and provided an overall correct classification rate of 88.7% for all five species. Classification scores ranged from a low of 50% (Sowerby's beaked whale) to a high of 100% (True's beaked whale, @fig-satlantic); all classification results were greater than expected (Priors in @fig-natlantic a). The sample sizes for the BANTER model were kept low (event sampsize = 1) to retain Sowerby's beaked whale, which had a sample size of 2. Analysis with a larger sample size resulted in improved classification results for Cuvier's and Gervais' beaked whale (see Supplement Fig. 3 for EC_IPI_ALT), but resulted in the loss of Sowerby's in the final model. Results from the EC_IPI model are presented here; results from the EC and EC_IPI_ALT model can be found in supplementary materials (Supplement Fig. 2, 3).

@@

The best model considered the echolocation pulses, IPI, and environmental features (EC_IPI_ENV Model, Detector Model: *sampsize* = 1, *ntree* = 10,000; Event model: *sampsize* = 5, *ntree* = 10,000) and provided an overall correct classification rate of 91.9% for all six species. Classification scores ranged from a low of 91.2% (BW43) to 100% for Cross Seamount beaked whale (@fig-epacific a). All classification results were greater than expected (Prior in @fig-epacific a). Results from the EC_IPI_ENV model are presented here; results from the EC and EC_IPI model can be found in supplementary materials (Supplement Figs. 5, 6).

@@

Previous application of BANTER to dolphin species in the California Current found that large sample sizes could result in strong classification of species where experienced analysts are unable to differentiate species (e.g., long-beaked and short-beaked common dolphins, @rankin_acoustic_2017). This suggests that large increases in sample sizes may improve classification results for Cuvier's beaked whales.

@@

These two examples of small sample sizes highlight a conflict in preferred protocol. Ideally, (1) all [species]{.underline} would be included in a classification model, so as to better represent the local species diversity, (2) all [events]{.underline} would ideally be included in the classification model, so as to better represent the variability found in the area, and (3) only confident 'ground truth' classifications would be considered in the training model. Unfortunately, in species where identity must be determined based on call characteristics (rather than visual confirmation of species identity in the field), it can sometimes be difficult to confidently determine species identity when calls are highly variable and do not include at least one call that provides high confidence of species identity. In the case of beaked whales, we recommend that training data include a high level of confidence for inclusion. This is especially critical for species with small sample sizes. An alternative is to require agreement from multiple analysts. This concern about accurate labeling of training data is further complicated by the potential existence of species that have not yet been identified. For example, [@barlow_unique_2022] recently discovered what appears to be a new species of beaked whale off Baja California, Mexico; this putative new species may have been detected but misclassified in other datasets (expert analysts did not identify these in the EPacific data presented here).

@@

The authors would like to acknowledge the large number of scientists and shipboard crew who were responsible for data collection during multiple large scale studies. Thanks to Liam Mueller-Brennan for their help with data analysis. Sea surface temperature data were provided by JPL under support by the NASA MEaSUREs program. Funding for data collection and analysis were provided by the U.S. Navy, the Bureau of Ocean Energy Management, and the National Oceanographic and Atmospheric Administration. The AMAPPS study (NAtlantic/SAtantic) was funded in part by the U.S. Department of the Interior, Bureau of Ocean Energy Management, Environmental Studies Program, Washington, DC, through Inter-Agency Agreement Number M10PG00075 with the National Marine Fisheries Service as the Atlantic Marine Assessment Program for Protected Species. The HICEAS 2017 Survey (Hawaii) was funded by Bureau of Ocean Energy Management under (IAA M17PG00024) and NMFS Pacific Islands Fisheries Science Center. The CCES Survey (EPacific) was funded by Bureau of Ocean Energy Management (No. M17PG00025), U.S. Navy Pacific Fleet Environmental Readiness Division (IAA No. N00070-18-MP-4C560) and Chief of Naval Operations N45, and Southwest Fisheries Science Center. The manuscript was improved thanks to reviews from Cory Hom-Weaver.
